plyranges: A grammar of genomic data transformation

### 2018-09-20 paper.tex
new pdf

\noindent The Bioconductor project provides many interoperable data abstractions
for analysing high-throughput genomics experiments; however implementing
a typical genomic workflow with Bioconductor requires learning these
abstractions and understanding them at an integrative level. This places
a large cognitive burden on the user, especially for non-programmers. To
reduce this burden we have created a grammar of genomic data
transformation that operates on a single, central Bioconductor data
structure, GRanges, which naturally represents genomic intervals and
their associated measurements. The grammar defines verbs for performing
actions on and between genomic interval data through a simplified,
coherent interface to existing Bioconductor infrastructure, resulting in
fluent analysis workflows. We have implemented this grammar as an
R/Bioconductor package called \texttt{plyranges}.

@@

Several domain specific languages (DSLs) enable scientists to process
and reason about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use a
SQL-like syntax for fast retrieval of information of unprocessed
sequencing data \autocites{Kozanitis2014-va}{Kozanitis2016-bm}.
Similarly, the Genometric Query Language (GMQL) implements a relational
algebra for combining genomic datasets \cite{Kaitoua2017-pw}. The
command line application BEDtools develops an extensive algebra for
performing arithmetic between two or more sets of genomic regions
\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be
evaluated either at the command line or embedded in scripts for batch
processing. They exist in a sparse ecosystem, mostly consisting of UNIX
and database tools that lack biological semantics and operate at the
level of files and database tables.

The Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
\autocites{r-core}{Lawrence2013-wg}{Huber2015-ei} define a DSL for
analysing genomics data with R, an interactive data analysis environment
that encourages reproducibility and provides high-level abstractions for
manipulating, modelling and plotting data, through state of the art
methods in statistical computing. The packages define object-oriented
(OO) abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs that are
embedded in programming languages include pybedtools and valr
\autocites{Dale2011-js}{Kent2017}, however these packages lack the
interoperability provided by the aforementioned Bioconductor packages
and are not easily extended.

@@

Recently, there has been a concerted, community effort to standardize R
data structures and workflows around the notion of tidy data
\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular data
structure that has observations as rows and columns as variables, and
all measurements pertain to a single observational unit. The tidy data
pattern is useful because it allows us to see how the data relate to the
design of an experiment and the variables measured. The \texttt{dplyr}
package \cite{Wickham2017-dplyr} defines an API that maps notions from
the general relational algebra to operations on tidy data. It expresses
each operation as a cohesive, endomorphic verb. Taken together these
features enable a user to write human readable analysis workflows.

@@

\hypertarget{genomic-relation-algebra}{%
\subsection{Genomic Relation Algebra}\label{genomic-relation-algebra}}

@@

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is essentially a constrained table, with fixed
columns for the chromosome, start and end coordinates, and the strand,
along with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment (figure
\ref{fig:GRanges}). GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic workflow,
while also being semantically rich enough to support high-level
operations on genomic ranges. As a core data structure, GRanges enables
interoperability between \texttt{plyranges} and the rest of
Bioconductor. Adhering to a single data structure simplifies the API and
makes it easier to learn and understand, in part because operations
become endomorphic, i.e., they return the same type as their input.

GRanges follows the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable describing the
observations. GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
dplyr. Thus, we define our algebra as an extension of the \texttt{dplyr}
algebra, and borrow its syntax conventions and design principles.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Arithmetic (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Arithmetic (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merging & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise union \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restriction & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorised into one of: aggregation, unary or binary
arithmetic, merging, modifier, or restriction. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

To support merging, our algebra recasts finding overlaps or nearest
neighbours between two genomic regions as variants of the relational
join operator. A join acts on two GRanges objects, a query and a
subject. The join operator is relational in the sense that metadata from
the query and subject ranges is retained in the joined range. All join
operators in the \texttt{plyranges} DSL generate a set of hits based on
overlap or proximity of ranges and use those hits to merge the two
datasets in different ways. There are four supported matching
algorithms: \emph{overlap}, \emph{nearest}, \emph{precede}, and
\emph{follow} (figures \ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can
further restrict the matching by whether the query is completely
\emph{within} the subject, and adding the \emph{directed} suffix ensures
that matching ranges have the same direction (strand).

{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-002}

\caption{Illustration of the three overlap join operators. Each join takes query and subject range as input (black and light gray rectangles, respectively). An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range. This index is used to expand the query ranges by where it was 'hit' by the subject ranges. The join semantics alter what is returned: for an \textbf{inner} join the query range is returned for each match, for an \textbf{intersect} the intersection is taken between overlapping ranges, and for a \textbf{left} join all query ranges are returned even if the subject range does not overlap them. This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
\end{figure}

For merging based on the hits, we have three modes: \emph{inner},
\emph{intersect} and \emph{left}. The \emph{inner} overlap join is
similar to the conventional inner join in that there is a row in the
result for every match. A major difference is that the matching is not
by identity, so we have to choose one of the ranges from each pair. We
always choose the left range. The \emph{intersect} join uses the
intersection instead of the left range. Finally, the overlap \emph{left}
join is akin to left outer join in Cobb's relational algebra: it
performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-003}

\caption{Illustration of neighbour finding joins. Each join takes a query and subject range and computes a 'Hits' object. For the \textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range. For the \textbf{follow} join there is only one query range that follows any subject range. Likewise for the \textbf{precede} join, there is only one query range that precedes a subject range.}\label{fig:nn-fig}
\end{figure}

Since the GRanges object is a tabular data structure, our grammar
includes operators to filter, sort and aggregate by columns in a
GRanges. These operations can be performed over partitions formed using
the \emph{group\_by} modifier. Together with our algebra for arithmetic
and merging, these operations conform to the semantics and syntax of the
\texttt{dplyr} grammar.

@@

Here we provide illustrative examples of using the \texttt{plyranges}
DSL to show how our grammar could be integrated into genomic data
workflows. We also highlight how interoperability with existing
Bioconductor infrastructure, enables easy access to public datasets and
methods for analysis and visualisation.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bigwig}\NormalTok{(}\DataTypeTok{overlap_ranges =}\NormalTok{ chr10_ranges) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

After filtering for regions with a coverage score greater than 8, we can
reduce individual runs to ranges representing the islands of coverage by
using the \texttt{reduce\_ranges()} function. This function allows a
summary to be computer over each island: in this case we take the
maximum of the scores to find the coverage peaks over chromosome 10.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{all_peaks <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(score }\OperatorTok{>}\StringTok{ }\DecValTok{8}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{reduce_ranges}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{max}\NormalTok{(score))}
\end{Highlighting}
\end{Shaded}

Returning to the GRanges object containing normalised coverage scores,
we filter to find the coordinates of the peak containing the maximum
coverage score. We can then find a 5000 nt region centered around the
maximum position by anchoring and modifying the width.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(chr10_max_score_region)}
\end{Highlighting}
\end{Shaded}

@@

We can directly obtain the genome information from the header of the BAM
file: in this case the reads were aligned to the hg19 genome build and
there are no reads overlapping the mitochondrial genome. To generate
bins of fixed width 10000nt we apply the \texttt{tile\_ranges()}
function to the genomic coordinates in locations.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{locations <-}\StringTok{ }\NormalTok{h1_bam_sorted }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bam}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }

\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{tile_ranges}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L)}
\end{Highlighting}
\end{Shaded}

Next we only read in alignments that overlap the genomic locations we
are interested in and select the query sequence. Note that the reading
of the BAM file is deferred: only alignments that pass the filter are
loaded into memory. We can add another column representing the GC
proportion for each alignment using the \texttt{letterFrequency()}
function from the \texttt{Biostrings} package \cite{R-biostrings}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{alignments <-}\StringTok{ }\NormalTok{h1_bam_sorted }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{read_bam}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter_by_overlaps}\NormalTok{(locations) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(seq) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{letterFrequency}\NormalTok{(seq, }\StringTok{"GC"}\NormalTok{, }\DataTypeTok{as.prob =} \OtherTok{TRUE}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

Finally, we use \texttt{group\_by\_overlaps()} to see where the
alignments overlap the genomic windows and then apply
\texttt{summarize()} to compute the total number of reads and average GC
content within each window.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{alignments_summary <-}\StringTok{ }\NormalTok{bins }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(alignments }\OperatorTok{%>%}\StringTok{ }\KeywordTok{select}\NormalTok{(score)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{(), }\DataTypeTok{avg_gc =} \KeywordTok{mean}\NormalTok{(score))}
\end{Highlighting}
\end{Shaded}

>>>>

\noindent The Bioconductor project provides many interoperable data abstractions
for analyzing high-throughput genomics experiments; however implementing
a typical genomic workflow with Bioconductor requires learning these
abstractions and understanding them at an integrative level. This places
a large cognitive burden on the user, especially for non-programmers. To
reduce this burden we have created a grammar of genomic data
transformation that operates on a single, central Bioconductor data
structure, GRanges, which naturally represents genomic intervals and
their associated measurements. The grammar defines verbs for performing
actions on and between genomic interval data through a simplified,
coherent interface to existing Bioconductor infrastructure, resulting in
fluent analysis workflows. We have implemented this grammar as an
R/Bioconductor package called \texttt{plyranges}.

@@

Several domain specific languages (DSLs) enable scientists to process
and reason about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use a
SQL-like syntax for fast retrieval of information of unprocessed
sequencing data \autocites{Kozanitis2014-va}{Kozanitis2016-bm}.
Similarly, the Genometric Query Language (GMQL) implements a DSL for
combining genomic datasets \cite{Kaitoua2017-pw}. The command line
application BEDtools develops an extensive algebra for performing
arithmetic between two or more sets of genomic regions
\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be
evaluated either at the command line or embedded in scripts for batch
processing. They exist in a sparse ecosystem, mostly consisting of UNIX
and database tools that lack biological semantics and operate at the
level of files and database tables.

The Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
\autocites{r-core}{Lawrence2013-wg}{Huber2015-ei} define a DSL for
analyzing genomics data with R, an interactive data analysis environment
that encourages reproducibility and provides high-level abstractions for
manipulating, modelling and plotting data, through state of the art
methods in statistical computing. The packages define object-oriented
(OO) abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs that are
embedded in programming languages include pybedtools and valr
\autocites{Dale2011-js}{Kent2017}, however these packages lack the
interoperability provided by the aforementioned Bioconductor packages
and are not easily extended.

@@

Recently, there has been a concerted, community effort to standardize R
data structures and workflows around the notion of tidy data
\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular data
structure that has observations as rows and columns as variables, and
all measurements pertain to a single observational unit. The tidy data
pattern is useful because it allows us to see how the data relate to the
design of an experiment and the variables measured. The \texttt{dplyr}
package \cite{Wickham2017-dplyr} defines an application programming
interface (API) that maps notions from the general relational algebra to
verbs that act on tidy data. These verbs can be composed together on one
or more tidy datasets with the pipe operator from the \texttt{magrittr}
package \cite{R-magrittr}. Taken together these features enable a user
to write human readable analysis workflows.

@@

\hypertarget{genomic-relational-algebra}{%
\subsection{Genomic Relational
Algebra}\label{genomic-relational-algebra}}

@@

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is a constrained table, with fixed columns for
the chromosome, start and end coordinates, and the strand, along with an
arbitrary set of additional columns, consisting of measurements or
metadata specific to the data type or experiment (figure
\ref{fig:GRanges}). GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic workflow,
while also being semantically rich enough to support high-level
operations on genomic ranges. As a core data structure, GRanges enables
interoperability between \texttt{plyranges} and the rest of
Bioconductor. Adhering to a single data structure simplifies the API and
makes it easier to learn and understand, in part because operations
become endomorphic, i.e., they return the same type as their input.

GRanges follow the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable describing the
observations. GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
\texttt{dplyr}. Thus, we define our algebra as an extension of the
\texttt{dplyr} algebra, and borrow its syntax conventions and design
principles.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merge & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise union \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \textit{x} that have been overlapped 
(the rectangles of \textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \textit{y} that overlap 
the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
ranges in \textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor
operations. The join semantics alter what is returned: \textbf{A}: for an 
\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
are returned. The returned ranges also include the metadata from the \textit{y}
range that overlapped the three \textit{x} ranges. \textbf{B} An 
\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \textit{x} ranges and the 
\textit{y} ranges. \textbf{C} For the \textbf{left} join all 
\textit{x} ranges are returned regardless of whether they are overlapped by 
\textit{y}. In this case the third range (rectangle with the asterisk next to it)
of the join would have missing values on metadata columns that came from \textit{y}.}
\label{fig:olaps-fig}
\end{figure}

To support merging, our algebra recasts finding overlaps or nearest
neighbors between two genomic regions as variants of the relational join
operator. A join acts on two GRanges objects: \emph{x} and \emph{y}. The
join operator is relational in the sense that metadata from the \emph{x}
and \emph{y} ranges are retained in the joined range. All join operators
in the \texttt{plyranges} DSL generate a set of hits based on overlap or
proximity of ranges and use those hits to merge the two datasets in
different ways. There are four supported matching algorithms:
\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}
(figure \ref{fig:olaps-fig}). We can further restrict the matching by
whether the query is completely \emph{within} the subject, and adding
the \emph{directed} suffix ensures that matching ranges have the same
direction (strand).

For merging based on the hits, we have three modes: \emph{inner},
\emph{intersect} and \emph{left}. The \emph{inner} overlap join is
similar to the conventional inner join in that there is a row in the
result for every match. A major difference is that the matching is not
by identity, so we have to choose one of the ranges from each pair. We
always choose the left range. The \emph{intersect} join uses the
intersection instead of the left range. Finally, the overlap \emph{left}
join is akin to left outer join in Codd's relational algebra: it
performs an overlap inner join but also returns all \emph{x} ranges that
are not hit by the \emph{y} ranges.

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
\caption{Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and 
\texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate 
operation that returns the same result. 
In each example, we have two BED files consisting of SNPs that are
genome-wide association study (GWAS) hits and reference exons. Each code block
counts for each SNP the number of distinct exons it overlaps. The 
\texttt{plyranges}code achieves this with an overlap join followed by partitioning
and aggregation. Strand is ignored by default here. 
The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' 
classes and their methods.}
\label{fig:code}
\end{figure}

Since the GRanges object is a tabular data structure, our grammar
includes operators to filter, sort and aggregate by columns in a
GRanges. These operations can be performed over partitions formed using
the \emph{group\_by} modifier. Together with our algebra for arithmetic
and merging, these operations conform to the semantics and syntax of the
\texttt{dplyr} grammar. Consequently, \texttt{plyranges} code is
generally more compact than the equivalent \texttt{GenomicRanges} code
(figure \ref{fig:code}).

@@

Here we provide illustrative examples of using the \texttt{plyranges}
DSL to show how our grammar could be integrated into genomic data
workflows. As we construct the workflows we show the data output
intermittently to assist the reader in understanding the pipeline steps.
The workflows highlight how interoperability with existing Bioconductor
infrastructure, enables easy access to public datasets and methods for
analysis and visualization.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bigwig}\NormalTok{(}\DataTypeTok{overlap_ranges =}\NormalTok{ chr10_ranges) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
\NormalTok{chr10_scores}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> GRanges object with 5789841 ranges and 1 metadata column:
#>             seqnames              ranges strand |              score
#>                <Rle>           <IRanges>  <Rle> |          <numeric>
#>         [1]    chr10             1-60602      * | 0.0422799997031689
#>         [2]    chr10         60603-60781      * |  0.163240000605583
#>         [3]    chr10         60782-60816      * |  0.372139990329742
#>         [4]    chr10         60817-60995      * |  0.163240000605583
#>         [5]    chr10         60996-61625      * | 0.0422799997031689
#>         ...      ...                 ...    ... .                ...
#>   [5789837]    chr10 135524723-135524734      * |  0.144319996237755
#>   [5789838]    chr10 135524735-135524775      * |  0.250230014324188
#>   [5789839]    chr10 135524776-135524784      * |  0.427789986133575
#>   [5789840]    chr10 135524785-135524806      * |  0.730019986629486
#>   [5789841]    chr10 135524807-135524837      * |   1.03103005886078
#>   -------
#>   seqinfo: 25 sequences from hg19 genome
\end{verbatim}

We then filter for regions with a coverage score greater than 8, and
following this reduce individual runs to ranges representing the islands
of coverage. This is achieved with the \texttt{reduce\_ranges()}
function, which allows a summary to be computed over each island: in
this case we take the maximum of the scores to find the coverage peaks
over chromosome 10.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{all_peaks <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(score }\OperatorTok{>}\StringTok{ }\DecValTok{8}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{reduce_ranges}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{max}\NormalTok{(score))}
\NormalTok{all_peaks}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> GRanges object with 1085 ranges and 1 metadata column:
#>          seqnames              ranges strand |            score
#>             <Rle>           <IRanges>  <Rle> |        <numeric>
#>      [1]    chr10     1299144-1299370      * | 13.2264003753662
#>      [2]    chr10     1778600-1778616      * | 8.20512008666992
#>      [3]    chr10     4613068-4613078      * | 8.76027011871338
#>      [4]    chr10     4613081-4613084      * | 8.43659973144531
#>      [5]    chr10             4613086      * | 8.11507987976074
#>      ...      ...                 ...    ... .              ...
#>   [1081]    chr10 135344482-135344488      * | 9.23237991333008
#>   [1082]    chr10 135344558-135344661      * |  11.843409538269
#>   [1083]    chr10 135344663-135344665      * | 8.26965999603271
#>   [1084]    chr10 135344670-135344674      * | 8.26965999603271
#>   [1085]    chr10 135345440-135345441      * | 8.26965999603271
#>   -------
#>   seqinfo: 25 sequences from hg19 genome
\end{verbatim}

Returning to the GRanges object containing normalized coverage scores,
we filter to find the coordinates of the peak containing the maximum
coverage score. We can then find a 5000 nt region centered around the
maximum position by anchoring and modifying the width.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(chr10_max_score_region)}
\NormalTok{peak_region}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> GRanges object with 890 ranges and 2 metadata columns:
#>         seqnames            ranges strand |            score.x
#>            <Rle>         <IRanges>  <Rle> |          <numeric>
#>     [1]    chr10 21805891-21805988      * | 0.0206599999219179
#>     [2]    chr10 21805989-21806000      * | 0.0211200006306171
#>     [3]    chr10 21806001-21806044      * |  0.022069999948144
#>     [4]    chr10 21806045-21806049      * | 0.0215900000184774
#>     [5]    chr10 21806050-21806081      * | 0.0211200006306171
#>     ...      ...               ...    ... .                ...
#>   [886]    chr10          21810878      * |   5.24951982498169
#>   [887]    chr10          21810879      * |   5.83534002304077
#>   [888]    chr10 21810880-21810884      * |   6.44267988204956
#>   [889]    chr10 21810885-21810895      * |   7.07054996490479
#>   [890]    chr10 21810896-21810911      * |   6.44267988204956
#>                  score.y
#>                <numeric>
#>     [1] 29.9573001861572
#>     [2] 29.9573001861572
#>     [3] 29.9573001861572
#>     [4] 29.9573001861572
#>     [5] 29.9573001861572
#>     ...              ...
#>   [886] 29.9573001861572
#>   [887] 29.9573001861572
#>   [888] 29.9573001861572
#>   [889] 29.9573001861572
#>   [890] 29.9573001861572
#>   -------
#>   seqinfo: 25 sequences from hg19 genome
\end{verbatim}

@@

We can directly obtain the genome information from the header of the BAM
file: in this case the reads were aligned to the hg19 genome build and
there are no reads overlapping the mitochondrial genome.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{locations <-}\StringTok{ }\NormalTok{h1_bam_sorted }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bam}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }
\end{Highlighting}
\end{Shaded}

Next we only read in alignments that overlap the genomic locations we
are interested in and select the query sequence. Note that the reading
of the BAM file is deferred: only alignments that pass the filter are
loaded into memory. We can add another column representing the GC
proportion for each alignment using the \texttt{letterFrequency()}
function from the \texttt{Biostrings} package \cite{R-biostrings}. After
computing the GC proportion as the score column, we drop all other
columns in the GRanges object.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{alignments <-}\StringTok{ }\NormalTok{h1_bam_sorted }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{read_bam}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter_by_overlaps}\NormalTok{(locations) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(seq) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{score =} \KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{letterFrequency}\NormalTok{(seq, }\StringTok{"GC"}\NormalTok{, }\DataTypeTok{as.prob =} \OtherTok{TRUE}\NormalTok{))}
\NormalTok{  ) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(score)}
\NormalTok{alignments}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> GRanges object with 8275595 ranges and 1 metadata column:
#>             seqnames            ranges strand |             score
#>                <Rle>         <IRanges>  <Rle> |         <numeric>
#>         [1]    chr10       50044-50119      - | 0.276315789473684
#>         [2]    chr10       50050-50119      + |              0.25
#>         [3]    chr10       50141-50213      - | 0.447368421052632
#>         [4]    chr10       50203-50278      + | 0.263157894736842
#>         [5]    chr10       50616-50690      + | 0.276315789473684
#>         ...      ...               ...    ... .               ...
#>   [8275591]     chrY 57772745-57772805      - | 0.513157894736842
#>   [8275592]     chrY 57772751-57772800      + | 0.526315789473684
#>   [8275593]     chrY 57772767-57772820      + | 0.565789473684211
#>   [8275594]     chrY 57772812-57772845      + |              0.25
#>   [8275595]     chrY 57772858-57772912      + | 0.592105263157895
#>   -------
#>   seqinfo: 24 sequences from an unspecified genome
\end{verbatim}

Finally, we create 10000nt tiles over the genome and compute the number
of reads and average GC content over all reads that fall within each
tile using an overlap join and merging endpoints.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{tile_ranges}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L)}

\NormalTok{alignments_summary <-}\StringTok{ }\NormalTok{bins }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(alignments) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{disjoin_ranges}\NormalTok{(}\DataTypeTok{n =} \KeywordTok{n}\NormalTok{(), }\DataTypeTok{avg_gc =} \KeywordTok{mean}\NormalTok{(score))}
\NormalTok{alignments_summary}
\end{Highlighting}
\end{Shaded}
